---
title: "CS555 - Assignment-3"
author: "Neha Bais"
date: "10/9/2020"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

>
The data in this document gives the number of meals eaten that contain fish (per week) and mercury levels in head hair for 100 fisherman.  Save the data to a format that can be read into R. Read the data in for analysis.  
Use R to calculate the quantities and generate the visual summaries requested below.  

```{r}
meal.data <- read.csv('meal_data.csv')
head(meal.data)
``` 

###### 1. To get a sense of the data, generate a scatterplot (using an appropriate window, label the axes, and title the graph).  Consciously decide which variable should be on the x-axis and which should be on the y-axis. Using the scatterplot, describe the form, direction, and strength of the association between the variables.   

```{r}    

attach(meal.data)

# Scatter plot
plot(Meals.with.fish, Total.Mercury, pch = 16, cex = 1, col = "red",
     main = "Mercury content vs Number of fish meal", 
     xlab = "No. of meals with Fish", ylab = "Total Mercury (mg/g)")
```  

>1. In this data, number of meals containing fish is describing the mercury content in head hair for the fisherman.  
***Explanatory Variable***: Meals.with.Fish plotted on x-axis.
***Response variable***: Total.Mercury will be plotted on y-axis. 
2. ***Form***: The points are following a linear pattern.
3. ***Direction***: Positively associated, as the increase in number of fish meal leads to increase in mercury level in fisherman.
4. ***Strength***: Moderately strong positive association as the points don't lie perfectly along a straight line.      
  
###### 2. Calculate the correlation coefficient.  What does the correlation tell us? 
```{r}

correlation <- cor(Meals.with.fish, Total.Mercury)
print(paste('Correlation between two: ',round(correlation,4)))

```
>Since correlation coefficient is positive, it indicates that the two values are positively correlated.
However the correlation between them is not that strong and can be described as moderately strong association.  
  
###### 3. Find the equation of the least squares regression equation, and write out the equation.  Add the regression line to the scatterplot you generated above.  
  
```{r}

my.model <- lm(Total.Mercury ~ Meals.with.fish)
print(my.model)

```

> *Equation of regression line*:   $Total.Mercury = 1.688 + 0.276 * Meals.with.fish$ 
                               

Plotting the line

```{r}

# Scatter plot
plot(Meals.with.fish, Total.Mercury, pch = 16, cex = 1, col = "red",
     main = "Mercury content vs Number of fish meal", 
     xlab = "No. of meals with Fish", ylab = "Total Mercury (mg/g)")
abline(my.model, col = 'blue')

```
  
###### 4. What is the estimate for beta1?  How can we interpret this value?  What is the estimate for beta0?  What is the interpretation of this value?  
  
> 1. Beta1 gives the predicted change in the response variable for one unit increase in explanatory variable. It also gives insight on the direction of reationship between the two variables.
2. For our data, beta1 = 0.276, this can be interpreted as the increase in mercury level in head hair of fisherman for every one additional meal containing fish. That is, for each additional fish meal the fisherman takes every week, the mercury level in their head hair improve by around 0.27% on average.
3. The estimate for beta0 is meaningful in this case. Here, beta0 = 1.688 which indicates that, on an average the mercury level in head hair of fisherman who does not take any meal having fish is around 1.68.  
  
  
###### 5. Calculate the ANOVA table and the table which gives the standard error of beta1.  Formally test the hypothesis that beta1 = 0 using either the F -test or the  t-test at the  alpha = 0.10 level.  Either way, present your results using the 5 step procedure as in the course notes.  
  
```{r anova table}

anova.table <- data.frame(anova(my.model))
print(anova.table)

```

Summary table for standard error of beta1.
```{r Std Error for beta1}

summary(my.model)
```
> Standard error for beta1 = 0.02851.  
  
### Hypothesis Test  

> 1. Set up the hypothesis and select the alpha level
        + H0 : beta1 = 0 (There is no linear association)
        + H1 : beta1 ≠ 0 (There is a linear association)
        + alpha = 0.10  
        
> 2. Select the appropriate test statistic 

$F = (Reg Mean square) / (Residual Mean square)$  with df = 1 and n-2 

> 3. State the decision rule
        + *Decision Rule*: Rejecth0 if p <= alpha (0.10) or F >= F.critical (2.75743)
        + Otherwise, do not reject H0  

> 4. Compute the test statistic and the associated p-value
        + From the anova table, F value = 93.688 and $p-value = 6.013e-16$

> 5. ***Conclusion***: 
        + Reject H0 since 93.688 > 2.75743. Also, p-value < 0.10
        + We have significant evidence at alpha = 0.10 that beta1 ≠ 0.
        + There is evidence of significant linear association between number of meal having fish and mercury level in head hair of fisherman. 
        
```{r}

F.critical <- qf(0.90, df1 = 1, df2 = nrow(meal.data)-2)
print(F.critical)

```
  
>calculate the  R-squared value and interpret this.  Also, calculate and interpret the 90% confidence interval for beta1.

***R-squared value****  
> + From summary table, we got Adjusted R-squared as 0.4835. This can be interpreted as, 48.35% of variability in mercusry level in head hair of fisherman can be explained by the number of fish meal they take per week.

***90% COnfidence Interval***


```{r}
confint(my.model, level = 0.90)

```
> 
+ 90% Confidence Interal = [0.228 ~ 0.323]
+ We can say that, if the fisherman increase their fish meal by 1 meal per week, the mercury level in their head hair increases between 0.22 and 0.32 mg/g.
  